---
title: 'Subsetting'
author: "D-Lab"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r include = FALSE}

# p_load loads and, if necessary, install missing packages.
# install.packages() + library() = p_load()
# If you just want to install, then use p_install()

if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, # for the tidyverse framework
  palmerpenguins,
  gapminder, 
  kableExtra, 
  flextable,
  modelr,
  nycflights13 
  )

```


# Subset Observations (Rows)

```{r}

penguins <- palmerpenguins::penguins
penguins
```

## Choose row by logical condition 

- Single condition

```{r}

penguins %>%
  filter(species == "Adelie") %>%
  arrange(desc(bill_length_mm))

```

The following filtering example was inspired by [the suzanbert's dplyr blog post](https://suzan.rbind.io/2018/02/dplyr-tutorial-3/).

- Multiple conditions (numeric)

```{r}

# First example
penguins %>%
  filter(bill_length_mm <= 42, bill_length_mm > 35) %>%
  nrow()
```


```{r}
# Same as above
penguins %>%
  filter(bill_length_mm <= 42 & bill_length_mm > 35) %>%
  nrow()
```


```{r}
# Not same as above
penguins %>%
  filter(bill_length_mm < 42 | bill_length_mm > 35) %>%
  nrow()

```

**Challenge 1**

1. (1) Use `filter(between())` to find characters whose heights are between 35.1 and 42.2 and (2) count the number of these observations.

- Multiple conditions (character)

```{r}
# Filter names include ars; `grepl` is a base R function  

penguins %>%
  filter(grepl("ent", tolower(species)))

```


```{r}
# Or, if you prefer dplyr way 

penguins %>%
  filter(str_detect(tolower(species), "ent"))

# Filter brown and black hair_color

penguins %>%
  filter(island %in% c("Biscoe", "Dream"))

```

**Challenge 2**

Use `str_detect()` to find all the penguins whose species includes "Chin"

```{r}

penguins %>%
  filter(str_detect(species, "Chin"))

```

## Choose row by position (row index)

```{r}

penguins %>%
  arrange(desc(bill_length_mm)) %>%
  slice(1:6)

```

## Sample by fraction


```{r}

# For reproducibility 
set.seed(1234)

# Old way 
penguins %>%
  sample_frac(0.10, 
              replace = FALSE) # Without replacement

# New way
set.seed(1234)
penguins %>%
  slice_sample(prop = 0.10, 
             replace = FALSE)

```

## Sample by number 

```{r}

# Old way 
set.seed(1234)
penguins %>%
  sample_n(20, 
           replace = FALSE) # Without replacement 

# New way
set.seed(1234)
penguins %>%
  slice_sample(n = 20, 
             replace = FALSE) # Without replacement 

```

## Top 10 rows orderd by height

```{r}
# Old way 
penguins %>%
  top_n(10, bill_length_mm) 

# New way
penguins %>%
  slice_max(bill_length_mm, n = 10) # Variable first, Argument second 

```

# Subset Variables (Columns)

```{r}

names(penguins)

```

### Select only numeric columns 

```{r}

# Only numeric 
penguins %>%
  select(where(is.numeric)) 

```

**Challenge 3** 

Use `select(where())` to find only non-numeric columns 

## Select the columns that include "bill" in their names 
 
```{r}

penguins %>%
  select(contains("bill"))

```

## Select the columns that include either "bill" or "mm" in their names 

- Basic R way 

`grepl` is one of the R base pattern matching functions. 

```{r}

penguins[grepl('bill|mm', names(penguins))]

```

**Challenge 4**

Use `select(match())` to find columns whose names include either "bill" or "mm". 

## Select the columns that starts with "b"

```{r}

penguins %>%
  select(starts_with("b"))

```

## Select the columns that ends with "mm"

```{r}

penguins %>%
  select(ends_with("mm"))

```

## Select the columns using both beginning and end string patterns 

The key idea is you can use Boolean operators (`!`, `&`, `|`)to combine different string pattern matching statements. 

```{r}

penguins %>%
  select(starts_with("b") & ends_with("mm"))

```

## Select bill_length_mm and move it before everything 

```{r}

# By specifying a column 
penguins %>%
  select(bill_length_mm, everything())
  
```

## Select variables from a character vector.

```{r}

penguins %>%
  select(any_of(c("species", "bill_length"))) %>%
  colnames()

```

## Select the variables named in the character + number pattern

```{r}
## This uses common dplyr functions to make a new variable. 
penguins_example <- penguins %>% 
  mutate(obs1 = 1,
         obs2 = 2,
         obs3 = 3,
         obs4 = 4,
         obs5 = 5)

penguins_example %>%
  select(num_range("obs", c(1:4)))

```

