---
title: 'Subsetting'
author: "Jae Yeon Kim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    number_sections: true
    toc: yes
  pdf_document:
    toc: yes
---

This workshop is for students who alreay have some experience with `dplyr` and `tidyr` and upgrade their data wrangling skills in R.

# Setup 

`pacman` is a great package management tool in R. (For more information, see [the package vignette](http://trinker.github.io/pacman/vignettes/Introduction_to_pacman.html).)

```{r}

# p_load loads and, if necessary, install missing pac￣kages. 
# install.packages() + library() = p_load()
# If you just want to install, then use p_install()

if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, # for the tidyverse framework
  gapminder # toy data
  )

```


# Subset Observations (Rows)

## Choose row by logical condition 

- Single condition 

```{r}

starwars %>%
  filter(gender == "female") %>%
  arrange(desc(height))

```

The following filtering example was inspired by [the suzanbert's dplyr blog post](https://suzan.rbind.io/2018/02/dplyr-tutorial-3/).

- Multiple conditions (numeric)

```{r}

# First example
starwars %>%
  filter(height < 180, height > 160) %>%
  nrow()

# Same as above
starwars %>%
  filter(height < 180 & height > 160) %>%
  nrow()

# Not same as above
starwars %>%
  filter(height < 180 | height > 160) %>%
  nrow()

# Same outcome as above but note the difference in boundary values
starwars %>%
  filter(between(height, 161, 179)) %>%
  nrow()

```

- Multiple conditions (character)

```{r}

# Filter names incldue ars; `grepl` is a base R function  

starwars %>%
  filter(grepl("ars", tolower(name)))

# Or, if you prefer dplyr way 

starwars %>%
  filter(str_detect(tolower(name), "ars"))

# Filter brown and black hair_color

starwars %>%
  filter(hair_color %in% c("black", "brown"))

```

## Choose row by position (row index)

```{r}

starwars %>%
  arrange(desc(height)) %>%
  slice(1:6)

```

## Sample by fraction

```{r}

# For reproducibility 
set.seed(1234)

# Extract 10% of rows 
starwars %>%
  sample_frac(0.10, 
              replace = FALSE) # Without replacement 

```

## Sample by number 

```{r}
# Extract 20 rows

starwars %>%
  sample_n(20, 
           replace = FALSE) # Without replacement 

```

## Top 10 rows orderd by height

```{r}

starwars %>% 
  top_n(10, height) 

```

# Subset Variables (Columns)

```{r}

names(msleep)

```

## Select the columns that include "sleep" in their names 
 
```{r}

msleep %>%
  select(contains("sleep"))

```

## Select the columns that include either "sleep" or "wt" in thier names 

- Basic R way 

`grepl` is one of the R base pattern matching functions. 

```{r}

msleep[grepl('sleep|wt', names(msleep))]

```

- dplyr 

`matches` does the same job in the `dplyr` universe. 

```{r}

msleep %>%
  select(matches("sleep_wt"))

```

## Select the columns that starts with "b"

```{r}

msleep %>%
  select(starts_with("b"))

```

## Select the columns that ends with "wt"

```{r}

msleep %>%
  select(ends_with("wt"))

```

## Select order and move it before everything 

```{r}

msleep %>%
  select(order, everything())

```

## Select the variables named as character 

```{r}

msleep %>%
  select(one_of(c("name", "order"))) %>%
  colnames()

```

## Select the variables named in the character + number pattern

```{r}

msleep$week8 <- NA

msleep$week12 <- NA

msleep %>%
  select(num_range("week", c(1:12)))

```
