---
title: 'Summarizing'
author: "Jae Yeon Kim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

This workshop is for students who already have some experience with `tidyr` and `dplyr` and `tidyr` and hope to upgrade their data wrangling skills in R.

# Setup

- `pacman` is a great package management tool in R (for more information, see [the package vignette](http://trinker.github.io/pacman/vignettes/Introduction_to_pacman.html)).

- Check your `dplyr` package is up-to-date by typing `packageVersion("dplyr")`. If the current installed version is less than 1.0, then update by typing `update.packages("dplyr")`. You may need to restart R to make it work.

```{r}

ifelse(packageVersion("dplyr") > 1, "The installed version of dplyr package is greater than or equal to 1.0.0", update.packages("dplyr"))

```

```{r}

# p_load loads and, if necessary, install missing packages. 
# install.packages() + library() = p_load()
# If you just want to install, then use p_install()

if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, # for the tidyverse framework
  gapminder, # toy example 
  kableExtra, 
  flextable 
  )

```

This part of the workshop heavily draws on [the suzanbert's dplyr blog post](https://suzan.rbind.io/2018/04/dplyr-tutorial-4/).

# Counting 

- How may countries in each continent?

```{r}

gapminder %>%
  count(continent)

```

- Let's arrange the result. 

```{r}

# Just add a new argument `sort = TRUE`
gapminder %>%
  count(continent, sort = TRUE)

# Same as above; How nice!
gapminder %>%
  count(continent) %>%
  arrange(desc(n))

```

**Challenge 1**
Count the number of observations per `continent` as well as `year` and arrange them with descending order. 

Let's take a deeper look at how things work under the hood.

- `tally()` works similar to `nrow()`: Calculate the total number of cases in a dataframe 
- `count` = `group_by()` + `tally()`

```{r}

gapminder %>%
  tally()

```

- `add_tally()` = `mutate(n = n())`

**Challenge 2** 

What does n in the below example represent? 

```{r}

gapminder %>%
  select(continent, country) %>%
  add_tally()

```

- `add_count`

Add count as a column 

```{r}

# Add count as a column
gapminder %>%
  group_by(continent) %>%
  add_count(year)

```

**Challenge 3**

Do the cases 1 and 2 in the below code chunk produce same outputs? If so, why?

```{r}

# Case 1
gapminder %>%
  group_by(continent, year) %>%
  count()

# Case 2
gapminder %>%
  group_by(continent) %>%
  count(year)

```

# Summarizing 

## Basic 

- Create a summary 

```{r}

gapminder %>%
  group_by(continent) %>%
  summarise(n = n(), 
            mean_gdp = mean(gdpPercap),
            sd_gdp = sd(gdpPercap))

tablea <- gapminder %>%
  group_by(continent) %>%
  summarise(n = n(), 
            mean_gdp = mean(gdpPercap),
            sd_gdp = sd(gdpPercap))

```
- Produce publishable tables 

```{r}

# For HTML and LaTeX
tablea %>% kableExtra::kable()

# For HTML and MS Office suite 
tablea %>% flextable::flextable()

```

## Conditional

### Old way 

- `summarise_all()`

```{r}

# Create a wide-shaped data example 
wide_gapminder <- gapminder %>%
  filter(continent == "Europe") %>%
  pivot_wider(names_from = country, 
              values_from = gdpPercap)

# Apply summarise_all 
wide_gapminder %>%
  select(-c(1:4)) %>%
  summarise_all(mean, na.rm = TRUE)

```

- `summarise_if()`: using a logical condition 

```{r}

wide_gapminder %>%
  summarise_if(is.double, mean, na.rm = TRUE)

```
- `summarise_at()`

- `vars() = select()`

```{r}

wide_gapminder %>%
  summarise_at(vars(-c(1:4)), 
               mean, na.rm = TRUE)

wide_gapminder %>%
  summarise_at(vars(contains("life")), 
               mean, na.rm = TRUE)

```
### New way 

- `summarise()` + `across()`

- If you find using `summarise_all()`, `summarise_if()` and `summarise_at()` confusing, here's a solution: use `summarise()` with `across()`.   

- `summarise_all()`

```{r}

wide_gapminder %>%
  summarise(across(Albania:`United Kingdom`, mean, na.rm = TRUE))

```
- `summarise_if()`

```{r}

wide_gapminder %>%
  summarise(across(is.double, mean, na.rm = TRUE))

```

- `summarise_at()`

```{r}

wide_gapminder %>%
  summarise(across(-c(1:4), 
               mean, na.rm = TRUE))

wide_gapminder %>%
  summarise(across(contains("life"), 
               mean, na.rm = TRUE))

```
Note that this workshop does not cover creating and manipulating variables using `mutate()` because many techniques you learned from playing with `summarise()` can be directly applied to `mutate()`. 

**Challenge 4**

1. Summarise average GDP of countries whose names starting with alphabet "A".

2. Turn the summary dataframe into a publishable table using either `kableExtra` or `flextable` package. 