---
title: 'Grouping and nesting'
author: "Jae Yeon Kim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

This workshop is for students who already have some experience with `tidyr` and `dplyr` and `tidyr` and hope to upgrade their data wrangling skills in R.

# Setup

- `pacman` is a great package management tool in R (for more information, see [the package vignette](http://trinker.github.io/pacman/vignettes/Introduction_to_pacman.html)).

- Check your `dplyr` package is up-to-date by typing `packageVersion("dplyr")`. If the current installed version is less than 1.0, then update by typing `update.packages("dplyr")`. You may need to restart R to make it work.

```{r}

ifelse(packageVersion("dplyr") > 1, "The installed version of dplyr package is greater than or equal to 1.0.0", update.packages("dplyr"))

```

```{r include = FALSE}

# p_load loads and, if necessary, install missing packages. 
# install.packages() + library() = p_load()
# If you just want to install, then use p_install()

if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, # for the tidyverse framework
  modelr, # for using the tidyverse framework to model outcomes 
  gapminder # data 
  )

```

# Grouping 

## Grouped summaries 

- Calculate the mean of `gdpPercap`.

```{r}

gapminder %>%
  group_by(continent) %>% # 
  summarise(mean_gdp = mean(gdpPercap))

```

- Calculate multiple summary statistics.

```{r}

gapminder %>%
  group_by(continent) %>% # 
  summarise(mean_gdp = mean(gdpPercap),
            count = n())

```

**Optional**

- Other summary statistics

1. Measures of spread: `median(x)`, `sd(x)`, `IQR(x)`, `mad(x)` (the median absolute deviation)

```{r}

# The Interquartile Range = The Difference Between 75t and 25t Percentiles 

gapminder %>%
  group_by(continent) %>% # 
  summarise(IQR_gdp = IQR(gdpPercap))

```

2. Measures of rank: `min(x)`, `quantile(x, 0.25)`, `max(x)`

```{r}

gapminder %>%
  group_by(continent) %>% # 
  summarise(min_gdp = min(gdpPercap),
            max_gdp = max(gdpPercap))

```

3. Measures of position: `first(x)`, `last(x)`, `nth(x, 2)`

```{r}

gapminder %>%
  group_by(continent) %>% 
  summarise(first_gdp = first(gdpPercap),
            last_gdp = last(gdpPercap))

gapminder %>%
  group_by(continent) %>% 
  arrange(gdpPercap) %>% # Adding arrange
  summarise(first_gdp = first(gdpPercap),
            last_gdp = last(gdpPercap))

```

4. Measures of counts: `n(x)` (all rows), `sum(!is.na(x))` (only non-missing rows) = `n_distinct(x)`

```{r}

gapminder %>%
  group_by(continent) %>%
  summarise(ns = n())

```

5. Counts and proportions of logical values: `sum(condition about x)` (the number of TRUEs in x), `mean(condition about x)` (the proportion of TRUEs in x)

```{r}

gapminder %>%
  group_by(continent) %>% 
  summarise(rich_countries = mean(gdpPercap > 20000))

```

# Nesting

## Nest

The following example comes from [R for Data Science](https://r4ds.had.co.nz/many-models.html) by by Garrett Grolemund and Hadley Wickham.

- How can you run multiple models simultaneously? Using a nested data frame. 

- **Grouped data: each row = an observation**

- **Nested data: each row = a group**

**Challenge 1**

In the following example, why did we use `country` and `continent` for nesting variables? 

```{r}

nested <- gapminder %>%
  group_by(country, continent) %>%
  nest() 

head(nested)

nested$data[[1]]

```

- Custom function 

```{r}

lm_model <- function(df) {
  
  lm(lifeExp ~ year, data = df)
}

```

- Apply function to the nested data 

```{r}

# Apply m_model to the nested data 

nested <- nested %>%
  mutate(models = map(data, lm_model)) # Add the list object as a new column

head(nested)

```
S3 is part of R's object oriented systems. If you need more information, check [this section](http://adv-r.had.co.nz/S3.html) in Hadley's Advanced R out.

## Unnest 

`glance()` function from `broom` package inspects the quality of a statistical model.

**Additional tips**

- `broom::glance(model)`: for evaluating model quality and/or complexity 
- `broom::tidy(model)`: for extracting each coefficient in the model (the estimates + its variability)
- `broom::augment(model, data)`: for getting extra values (residuals, and influence statistics)


```{r}

glanced <- nested %>%
  mutate(glance = map(models, broom::glance))

glanced$glance[[1]]

```

`unnest()` unpacks the list objects stored in glance column 

```{r}

glanced %>% 
  unnest(glance) %>%
  arrange(BIC) # Low to High; Lower BIC indicates a better model fit 

glanced %>% 
  unnest(glance) %>%
  ggplot(aes(continent, BIC)) +
    geom_jitter(width = 0.5)

```

# Mapping 

We tasted a little bit about how `map()` function works. Let's dig into it deeper as this family of functions is really useful. For more information, see Rebecca Barter's wonderful tutorial on the `purrr` package. In her words, this is "the tidyverse's answer to apply functions for iteration". `map()` function can take a vector (of any type), a list, and a dataframe for input. 

```{r}

multiply <- function(x){
  x*x 
}

df <- list(first_obs = rnorm(7, 1, sd =1),
           second_obs = rnorm(7, 2, sd = 2)) # normal distribution 

map(.x = df, 
    .f = multiply)
```
**Challenge 2**

Try `map_df(.x = df, .f = multiply)` and tell me what's the difference between the output you got and what you saw earlier. 

If you want to know more about the power and joy of functional programming in R (e.g., `purrr::map()`), then please take ["How to Automate Repeated Things in R"](https://github.com/dlab-berkeley/R-functional-programming) workshop.