---
title: 'Grouping and nesting'
author: "Jae Yeon Kim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    number_sections: true
    toc: yes
  pdf_document:
    toc: yes
---

This workshop is for students who already have some experience with `dplyr` and `tidyr` and hope to upgrade their data wrangling skills in R.

# Setup 

- `pacman` is a great package management tool in R. (For more information, see [the package vignette](http://trinker.github.io/pacman/vignettes/Introduction_to_pacman.html).)

- Check your `dplyr` package is up-to-date by typing `packageVersion("dplyr")`. If the current installed version is less than 1.0, then update by typing `update.packages("dplyr")`. You may need to restart R to make it work.

```{r include = FALSE}

# p_load loads and, if necessary, install missing packages. 
# install.packages() + library() = p_load()
# If you just want to install, then use p_install()

if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, # for the tidyverse framework
  modelr, # for using the tidyverse framework to model outcomes 
  gapminder # data 
  )

```

# Grouping 

## Grouped summaries 

- Calculate the mean of `gdpPercap`.

```{r}

gapminder %>%
  group_by(continent) %>% # 
  summarise(mean_gdp = mean(gdpPercap))

```

- Calculate multiple summary statistics.

```{r}

gapminder %>%
  group_by(continent) %>% # 
  summarise(mean_gdp = mean(gdpPercap),
            count = n())

```

- Other summary statistics

1. Measures of spread: `median(x)`, `sd(x)`, `IQR(x)`, `mad(x)` (the median absolute deviation)

```{r}

# The Interquartile Range = The Difference Between 75t and 25t Percentiles 

gapminder %>%
  group_by(continent) %>% # 
  summarise(IQR_gdp = IQR(gdpPercap))

```

2. Measures of rank: `min(x)`, `quantile(x, 0.25)`, `max(x)`

```{r}

gapminder %>%
  group_by(continent) %>% # 
  summarise(min_gdp = min(gdpPercap),
            max_gdp = max(gdpPercap))

```

3. Measures of position: `first(x)`, `last(x)`, `nth(x, 2)`

```{r}

gapminder %>%
  group_by(continent) %>% 
  summarise(first_gdp = first(gdpPercap),
            last_gdp = last(gdpPercap))

gapminder %>%
  group_by(continent) %>% 
  arrange(gdpPercap) %>% # Adding arrange
  summarise(first_gdp = first(gdpPercap),
            last_gdp = last(gdpPercap))

```

4. Measures of counts: `n(x)` (all rows), `sum(!is.na(x))` (only non-missing rows) = `n_distinct(x)`

```{r}

gapminder %>%
  group_by(continent) %>%
  summarise(ns = n())

```

5. Counts and proportions of logical values: `sum(condition about x)` (the number of TRUEs in x), `mean(condition about x)` (the proportion of TRUEs in x)

```{r}

gapminder %>%
  group_by(continent) %>% 
  summarise(rich_countries = mean(gdpPercap > 20000))

```


# Nesting

## Nest

The following example comes from [R for Data Science](https://r4ds.had.co.nz/many-models.html) by by Garrett Grolemund and Hadley Wickham.

- How can you run multiple models simultaneously? Using a nested data frame. 

- Grouped data: each row = an observation

- Nested data: each row = a group

```{r}

nested <- gapminder %>%
  group_by(country, continent) %>%
  nest() 

# country & continent (two grouping variables) + data
names(nested)

# Data = listed columns 

nested$data[[1]]

```

- Custom function 

```{r}

lm_model <- function(df) {
  
  lm(lifeExp ~ year, data = df)
}

```

- Apply function to the nested data 

```{r}

# Apply m_model to the nested data 

nested <- nested %>%
  mutate(models = purrr::map(data, lm_model)) # Add the list object as a new column

# Display the result by filtering rows 

nested %>%
  filter(continent == "Asia")

```

## Unnest 

`glance()` function from `broom` package inspects the quality of a statistical model.

**Additional tips**

- `broom::glance(model)`: for evaluating model quality and/or complexity 
- `broom::tidy(model)`: for extracting each coefficient in the model (the estimates + its variability)
- `broom::augment(model, data)`: for getting extra values (residuals, and influence statistics)


```{r}

glanced <- nested %>%
  mutate(glance = map(models, broom::glance))

```

`unnest()` unpacks the list objects stored in glance column 

```{r}

glanced %>% 
  unnest(glance) %>%
  arrange(BIC) # Low to High; Lower BIC indicates a better model fit 

glanced %>% 
  unnest(glance) %>%
  ggplot(aes(continent, BIC)) +
    geom_jitter(width = 0.5)

```

- Another use case of `unnest()` is for simplifying list columns. 

- The following example does not work because `quantitle(mpg)` has many objects. 

```{r eval=FALSE, include=FALSE}

mtcars %>%
  group_by(cyl) %>%
  summarise(q = quantile(mpg))

```

- Let's put these many objects in a list. Now, it works but it's difficult to know what's inside the listed-column `q`.

```{r}

mtcars %>%
  group_by(cyl) %>%
  summarise(q = list(quantile(mpg)))

```

- We can unpack the listed column using `unnest()`.

```{r}

mtcars %>%
  group_by(cyl) %>%
  summarise(q = list(quantile(mpg))) %>%
  unnest()

```