---
title: 'Tidy Data and Reshaping'
author: "Jae Yeon Kim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

**Challenge 1**: Why this data is not tidy?

```{r}

table4a

```

**Challenge 2**

1. Why this data is not tidy? (This exercise comes from [`pivot` function vigenette](https://tidyr.tidyverse.org/articles/pivot.html).) Too long or too wide?

```{r}

billboard

```

2. How can you fix it? Which pivot?

```{r}

# Old way
billboard %>%
  gather(key = "week",
         value = "rank",
         starts_with("wk")) %>% # Use regular expressions
  drop_na() # Drop NAs

```

- Note that `pivot_longer()` is more versatile than `gather()`.

```{r}

# New way
billboard %>%
  pivot_longer(
    cols = starts_with("wk"), # Use regular expressions
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE # Drop NAs
  )

```

**Challenge 3**

The example comes from [*R for Data Science*](https://r4ds.had.co.nz/tidy-data.html).

```{r}

stocks <- tibble(
  year = c(2019, 2019, 2019, 2020, 2020, 2020),
  qtr = c(1, 2, 3, 2, 3, 4), 
  return = c(1, 2, 3, NA, 2, 3)
)

```

- Where is explicit missing value?

- Does `stocks` have implicit missing values?

```{r}

# implicit missing values become explicit 
stocks %>%
  pivot_wider(names_from = year, 
              values_from = return)

```

**Challenge 4**

- This exercise comes from [`pivot` function vigenette](https://tidyr.tidyverse.org/articles/pivot.html).

- Could you make `station` a series of dummy variables using `pivot_wider()`?

```{r}
fish_encounters
```

1. Which pivot you should use?

```{r}
fish_encounters %>%
  pivot_wider(
    names_from = station, 
    values_from = seen 
  )
```

2. Are there explicit missing values? 

```{r}
fish_encounters %>%
  pivot_wider(
    names_from = station, 
    values_from = seen,
    values_fill = list(seen = 0)
  )
 
```

3. How could you turn these NAs into 0s? Check `values_fill` argument in the `pivot_wider()` function. 

