---
title: 'Subsetting'
author: "Jae Yeon Kim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

This workshop is for students who already have some experience with `tidyr` and `dplyr` and `tidyr` and hope to upgrade their data wrangling skills in R.

# Setup

- `pacman` is a great package management tool in R (for more information, see [the package vignette](http://trinker.github.io/pacman/vignettes/Introduction_to_pacman.html)).

- Check your `dplyr` package is up-to-date by typing `packageVersion("dplyr")`. If the current installed version is less than 1.0, then update by typing `update.packages("dplyr")`. You may need to restart R to make it work.

```{r}

ifelse(packageVersion("dplyr") > 1, 
       "The installed version of dplyr package is greater than or equal to 1.0.0",
       update.packages("dplyr"))

```

```{r}

# p_load loads and, if necessary, install missing packages. 
# install.packages() + library() = p_load()
# If you just want to install, then use p_install()

if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyverse, # for the tidyverse framework
  gapminder # toy data
  )

```


# Subset Observations (Rows)

## Choose row by logical condition 

- Single condition 

```{r}

starwars %>%
  filter(gender == "feminine") %>%
  arrange(desc(height))

```

The following filtering example was inspired by [the suzanbert's dplyr blog post](https://suzan.rbind.io/2018/02/dplyr-tutorial-3/).

- Multiple conditions (numeric)

```{r}

# First example
starwars %>%
  filter(height < 180, height > 160) %>%
  nrow() 

```

```{r}
# Same as above
starwars %>%
  filter(height < 180 & height > 160) %>%
  nrow()

```

```{r}
# Not same as above
starwars %>%
  filter(height < 180 | height > 160) %>%
  nrow()

```

**Challenge 1**

1. (1) Use `filter(between())` to find characters whose heights are between 160 and 180 and (2) count the number of these observations. (Don't include 160 and 180)

- Minimum reproducible example 

```{r}

df <- tibble(
  heights = c(160:180),
  char = rep("none", length(c(160:180)))
)

```

- Multiple conditions (character)

```{r}

# Filter names include ars; `grepl` is a base R function  

starwars %>%
  filter(grepl("ars", tolower(starwars$name)))

# Or, if you prefer dplyr way 

starwars %>%
  filter(str_detect(tolower(name), "ars"))

# Filter brown and black hair_color

starwars %>%
  filter(hair_color %in% c("black", "brown"))

```

**Challenge 2**

Use `str_detect()` to find characters whose names include "Han".

```{r}

starwars %>%
  filter(str_detect(name, "Han"))
  
```

## Choose row by position (row index)

```{r}

starwars %>%
  arrange(desc(height)) %>%
  slice(1:6)

```

## Sample by fraction

- In dplyr 1.0.0, `sample_frac(),` `sample_n(),` `top_n()` -> `slice_()` 

```{r}

# For reproducibility 
set.seed(1234)

# Old way 

starwars %>%
  sample_frac(0.10, 
              replace = FALSE) # Without replacement 

```

```{r}
# New way

starwars %>%
  slice_sample(prop = 0.10, 
             replace = FALSE)

```

## Sample by number 

```{r}

# Old way 

starwars %>%
  sample_n(20, 
           replace = FALSE) # Without replacement 

# New way

starwars %>%
  slice_sample(n = 20, 
             replace = FALSE) # Without replacement 

```

## Top 10 rows orderd by height

```{r}

# Old way 
starwars %>% 
  top_n(10, height) 

# New way
starwars %>%
  slice_max(height, n = 10) # Variable first, Argument second 

```

# Subset Variables (Columns)

```{r}

names(msleep)

glimpse(msleep)

```

### Select only numeric columns 

In dplyr 1.0.0., `select_all(),` `select_if(),` and `select_at()` -> `select(where())` and 

`mutate_all(),` `mutate_if(),` and `mutate_at()` -> `mutate(across())`

```{r}

# Only numeric 
msleep %>%
  select(where(is.numeric))

# Turn numeric into factor then select only factor 
msleep %>%
  mutate(across(is.numeric, as.factor)) %>%
  select(where(is.factor))

```

**Challenge 3** 

Use `select(where())` to find only non-numeric columns 

## Select the columns that include "sleep" in their names 
 
```{r}

msleep %>%
  select(contains("sleep"))

```

## Select the columns that include either "sleep" or "wt" in thier names 

- Basic R way 

`grepl` is one of the R base pattern matching functions. 

```{r}

msleep[grepl('sleep|wt', names(msleep))]

msleep %>%
  select(matches("sleep|wt"))

```

**Challenge 4**

Use `select(match())` to find columns whose names include either "sleep" or "wt". 

## Select the columns that starts with "b"

```{r}

msleep %>%
  select(starts_with("b"))

```

## Select the columns that ends with "wt"

```{r}

msleep %>%
  select(ends_with("wt"))

```

## Select the columns using both beginning and end string patterns 

The key idea is you can use Boolean operators (`!`, `&`, `|`)to combine different string pattern matching statements. 

```{r}

msleep %>%
  select(starts_with("b") & ends_with("wt"))

```

## Select order and move it before everything 

```{r}

# By specifying a column 
msleep %>%
  select(order, everything())
  
```

## Select variables from a character vector.

```{r}

msleep %>%
  select(any_of(c("name", "order"))) %>%
  colnames()

```

## Select the variables named in the character + number pattern

```{r}

msleep$week15 <- NA

msleep$week8 <- NA

msleep$week12 <- NA

msleep$week_extra <- 0 

msleep %>%
  select(num_range("week", c(1:10)))

```

- Change `c(1:10)` to `c(1:15)`. 

- Can you tell the difference?